$\XX$ dubious ways of estimating $\rr$

### 2018-07-03 paper/draft_v6.tex
finally on git

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
%\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{graphicx}% http://ctan.org/pkg/graphicx
\usepackage{systeme}
%% tables
\usepackage{booktabs}
\usepackage[table,xcdraw]{xcolor}
\usepackage{float}

@@

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'', $\mathcal{R}_0$, the reproduction number (by convention pronounced ``R-naught''), for nearly a century, has been an elusive quantity for epidemic modelers to estimate \citep{Heesterbeek2002}.  Defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  $\rr$, in some ways, summarizes an entire outbreak of a disease; it  is used to assess whether a disease will break out and what percentage of the population needs to be vaccinated to avoid such an outbreak.  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  The primary issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population.

@@

Additionally, we describe  interval estimates for the above methods for estimating $\rr$ in order to better answer the question of whether $\rr > 1$, which implies an outbreak of the disease.  These methods are the delta method and posterior distribution.

@@

The first set of methods are those of the survival function, which is closely  tied to the history of $\rr$.  The survival function, which originates from the field of demography in the late 1800s, describes how many female offspring a woman would produce \citep{dietz1993estimation}.  In demography, we have
\begin{align*}
\rr = \int_0^\infty p(a) \beta(a) da
\end{align*}
where $p(a)$ denotes the probability of a woman surviving to age $a$ and $\beta(a)$ the rate of an individual of age $a$ giving birth to a  girl.   This interpretation also explains the origins of the name of $\rr$, which was imported to the field of epidemiology by MacDonald and Smith \citep{dietz1993estimation}.  Analogously in epidemiology, $p(a)$ is the age of a disease, and $\beta$ is the transmission rate.

This is the chronologically first, and in many ways, the most direct method to estimate $\rr$.  However, this model also seems to be the most difficult to estimate as both $p(a)$ and $\beta (a)$ are random variables.   In general, it is also difficult to estimate these parameters before or even during an ongoing disease outbreak.  As such, $\rr$ can only be estimated after an outbreak, which means it cannot be used to determine ongoing prevention strategies.  However, having a post-disease estimation of $\rr$ may be useful for preventing future outbreaks of the disease and aids in comparing the severity of many diseases.  Another advantage of this method is that it is intuitive and independent of assumptions of how the disease is transmitted from person to person.    We describe four methods which estimate $\rr$ from the survival function:  the survival function, direct estimation, and direct parameter estimation.

@@

Direct parameter estimation, an approach for parameters that are readily available to be estimated, is a `plug-in' estimator approach as described by \cite{lipsitch2003} and \cite{dietz1993estimation}.  This method, like the survival function, is highly interpretable.  We assume that we can estimate $\rr$ if we know how many others an individual comes into contact $(k)$, the probability of transmission given contact $(b)$, and how long the disease lasts within an individual $(D)$.  Then $\rr$ is the product of these parameters, 
\begin{align}\label{eq:dpe}
\rr = kbD.
\end{align}
\cite{lipsitch2003} note that directly estimating $\rr$ in such a way may be difficult due to the available data, as in the previous  method.  They instead estimate $\rr$ using the mean serial interval, $\omega$, the time from onset of symptoms in an index case to the onset of symptoms in a subsequent case infected by the index patient in addition to the assumption of exponential growth of infected cases, as a proxy for the duration $D$.  An advantage of this method is that it reduces the problem to three parameters that we can conceivably collect data on.  A disadvantage is that to estimate $\rr$ alone,  data is only used from the initial outbreak of the disease, so often we have a small sample size, resulting in large confidence sets or credible intervals.

@@

A derivation of $\rr$ similar to the survival function comes from the idea of the von Foerster equations, which were used to calculate the number of cells in a ``deme,'' a constrained area.  The function $Y(t, \tau)$ calculates the number of cells of age $\tau$ at time $t$ \citep{trucco1965}.  In epidemiology, $Y$ represents the number of infections instead of cells. Letting $Y(t, \tau)$ be the number of infections at time $t$ of age $\tau$, the von Foerster equation is
\begin{align*}
  \frac{\partial Y(t, \tau)}{\partial t} +   \frac{\partial Y(t, \tau)}{\partial \tau} = -  \lambda \cdot n(t, a) ,
\end{align*}
where $Y(t,\tau)$ is the number of people at time $t$ who were infected time $\tau$ ago and $\lambda$ is the loss function, which we assume to be 0, that is no deaths.  The right hand side represents the net loss of cells.  This derivation is based on the ``reasoning that the number of cells which are in the age category $(a, a + \Delta a)$ at time $t$, is equal to $n(t,a) \Delta a$'' \citep{trucco1965}.     \cite{fraser2004factors} derived an expression for $\rr$ from the von Foerster equation and the additional boundary equation
\begin{align*}
  Y(t,0) &= \int_0^t \beta( \tau) Y(t, \tau) d \tau,
\end{align*}
with initial conditions $Y(0,0) = Y_0$ and $Y(t, \tau) = 0$ when $t < \tau$.  Then the reproduction number is given by Equation \ref{eq:vfr0},
\begin{align}\label{eq:vfr0}
\rr &= \int_0^\infty \beta(\tau) d\tau,
\end{align}
 where $\beta (\tau)$ represents the infectiousness at time $\tau$ since the infection.    Again, this model is more mathematical in nature than practical as $\beta (\tau)$ is difficult to estimate.  To estimate $\rr$, \cite{fraser2004factors} make an approximation that $Y(t, \tau) = K(\tau) \exp (rt)$ and further adapt the method to include prevention strategies.  The main function to be estimated is $\beta (\tau)$.  Estimating $\beta (\tau)$ has many of the same issues as trying to estimate $b(a)$ in the survival function in that it is difficult to estimate while the disease is ongoing but is also advantageous in that it does not place assumptions on how the disease is transmitted throughout the population.

@@

First introduced by \cite{Kermack700}, a large and common set of methods used estimate $\rr$ is that of fitting compartment models (CMs) to disease data.  As described in \cite{daley2001epidemic},  CMs model how objects move from one discrete state to another over time.  These models provide an aggregate model of how diseases may be spread throughout a population.  Within the CM framework, we make four essential assumptions,
\begin{enumerate}
\item The compartments are discrete and have no overlap.
\item The transition of objects into and out of compartments is described by a set of known equations, possibly dependent on unknown parameters.
\item The populations mix homogeneously
  \item The number of objects in each compartment at time $t=0$ is known
  \end{enumerate}

@@

In this section, we discuss $15$ different CM-based methods used to estimate $\rr$.  Different model specifications of CMs yield different mathematical derivations of $\rr$.  That is, for example, an SIR model with non-constant population will yield a different mathematical formula for $\rr$ than the SEIR model (here ``E'' stands for exposed) with non-constant population.  The difference in these two models is that the SEIR model has an extra compartment which in turn effects the estimates of transitions from one compartment to another.  That said, even within the same compartment model framework, we can estimate $\rr$ in a variety of ways using statistical techniques.  In fact, we describe $\xxsir$ methods to estimate $\rr$, within the SIR framework alone!  Most of these methods are recipes to estimate parameters within the SIR model and may be extended to other CM frameworks such as the SEIR model.  The final method described in this section, the next generation operator (NGM) is a general method to derive $\rr$ from any valid CM.  A particular advantage to CMs compared to survival function methods is that they are easier to estimate as the data required is typically collected about diseases: namely, incidence counts.

@@

In words, susceptible individuals become infected at a rate that is proportional to the percentage of infected individuals multiplied by $\beta$, the infection rate, and the number of susceptible individuals.  Infected individuals recover at a rate of $\gamma$ multiplied by the number of infected individuals.

An outbreak occurs if the rate of change of infected individuals is positive,
\begin{align*}
  \frac{dY}{dt} &> 0 \\
  \frac{\beta X Y}{N}  - \gamma Y &> 0 ,\\
  Y \left ( \beta \frac{X}{N} - \gamma \right ) & > 0
\end{align*}
that is,  the rate of new infections is greater than the rate of recovery.  So as long as the number of susceptibles is large, $\frac{X}{N} \approx 1$, then an outbreak will occur if $\rr >1$,
\begin{align}\label{eq:deriv_sir}
  \rr = \frac{\beta}{\gamma} > 1.
  \end{align}
Ideally, we have access data of the number of susceptibles, infecteds, and recovered individuals at different points over time.  From this we would like to estimate the $\beta$ and $\gamma$ parameters and ultimately $\rr$.

@@

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\text{argmin}_{\beta, \gamma} \sum_{t} \left [ \left (Y_{obs}(t) - Y(t)\right )^2 + \left ( X_{obs}(t) - X(t) \right )^2 \right ]
\end{align*}
Then  the estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{R}_0= \frac{\hat{\beta}}{\hat{\gamma}}.
  \end{align}
We can use gradient descent or another optimizer to find estimates for these parameters.  If we assume $Y= f(t) + \epsilon_t$ with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$ where $f(t) = \int_0^t \frac{dY}{ds} ds$, then this this method is equivalent to finding the maximum likelihood estimator (MLE) of $\beta$ and $\gamma$.  However, we note that the the $\epsilon_t$ are certainly not independent from one another and so valid inference is still an issue in this framework.  Even if it were correct, ratio estimatiors typically result in large, uninformative confidence intervals.   The advantage is that this is a very straightforward way to estimate $\rr$ from the SIR model.

@@

\begin{align*}
  \left \{
  \begin{array}{cl}
    \frac{dX}{dt} &= - \rr \gamma Y \frac{X}{N}\\
    \frac{dY}{dt} &=  \rr \gamma Y \frac{X}{N}  - \gamma Y \\
    \frac{dZ}{dt} &=  - \gamma Y 
  \end{array}
  \right .
  \end{align*}
We find
\begin{align*}
(\hat{R}_0, \hat{\gamma} ) &= \text{argmin}_{\beta, \gamma} \sum_{t} \left [ \left (Y_{obs}(t) - Y(t)\right )^2 + \left ( X_{obs}(t) - X(t) \right )^2 \right ]
\end{align*}
We use the \(\rr\) directly from the above estimation process, which again can be done with gradient descent or another optimization process.  Again, we cannot assume $\rr$ has a normal distribution as the errors are correlated.  This method is better than SIR LS for estimating $\rr$ as there is no ratio.

@@

Specifically, we fit two linear polynomials in \(t\) with degree \(K= 10\) to \(X_{obs}\)
and \(Y_{obs}\) using least squares to find the coefficients $\{(\hat{x}_k,
\hat{y}_k)\}_{k=1, \dots, K}$,
\begin{align*}
\hat{X}(t) &= \sum_{k=0}^K \hat{x}_k t^k\\
{\hat{Y}}(t) &= \sum_{k=0}^K \hat{y}_k t^k
\end{align*}
Then, we estimate the derivatives as
\begin{align*}
\hat{X}^\prime(t) &= \sum_{k=1}^K k \hat{x}_k t^{k-1}\\
\hat{Y}^\prime(t) &= \sum_{k=0}^K k \hat{y}_k t^{k-1}
\end{align*}
Following,  \(\rr\) is derived from the ODEs in Equation \eqref{eq::sir},
\begin{align}
  - \frac{X^\prime}{Z^\prime}&= \rr \frac{X}{N} \nonumber\\
\rr &=       \frac{X}{ X + Y} \cdot \frac{N}{X} \nonumber\\
  \hat{\rr} &= \frac{\hat{X}^\prime(0)}{ \hat{X}^\prime(0) + \hat{Y}^\prime(0)} \cdot \frac{N}{\hat{X}(0)}. \nonumber
  \end{align}
Like in the previous cases, we are assuming the true models are
\begin{align*}
X(t) &= \sum_{k=0}^K x_k t^k + \epsilon_t\\
  {Y}(t) &= \sum_{k=0}^K y_k t^k + \epsilon_t,
\end{align*}
with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$.  Here, $K=10$ is rather arbitrary and should be selected using some criterion such as AIC in addition with the knowledge of how many data points are available to the user.  Besides optionally deciding on the degree of polynomials to fit, this model is simple to implement and gives very comparable results to using least squares with the SIR model.  The time $t=0$ to best capture the initial outbreak.  Again, we cannot use this model for inference, as our error assumption is violated.  Another disadvantage is that we are using the first time point $t=0$ to estimate $\rr$.

@@

\subsubsection{Max of Data (SIR Max)}\label{max-of-data}
The I ($Y$) compartment of the SIR model has one peak.    We note that \(Y^\prime(t) = \frac{\beta XY}{N} - \gamma Y = 0\) when \(\rr = \frac{\beta}{\gamma} = \frac{N(t)}{X(t)}\). The max
of \(Y\) occurs when \(Y^\prime = 0\) and hence
\[\rr = \frac{N}{X_{obs}(t^*)},\] where
\(t^* = \text{arg max}_{t} Y_{obs}(t)\).   A major disadvantage to this method is that it only uses one sample, making it a highly variable statistic.  An advantage of this method is that it places no distributional assumptions on any parameter.

@@

Here we assume that the time step is small enough to approximate the incidence.  The advantage of this method is that incidence data is generally readily available as is prevalence data for certain diseases such as HIV.  However, as one is required to have prior knowledge about $\gamma$, it may be easier to directly estimate $\rr$ with one of the many other methods described that does not require a prior knowledge about $\gamma$.  Again, we are using only one time point to estimate $\rr$.

@@

In this framework, the number of susceptibles at the next step, $X_{t}$, has a Binomial distribution based on the contacts with the current number of infected, $Y_{t}$ and the current number of susceptibles.  That is $X_{t+1} \sim \text{Binom}(X_t, \alpha^{Y_t}$), where $\alpha$ is the probability of avoiding infection from an infective.  \cite{barbour2004} report that the reproduction number is thus,
\begin{align}\label{eq:r0-mc}
\rr &= \log \left ( \frac{1}{1-\alpha}\right ).
\end{align}

@@

Described by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, the sequential Bayes method is a Bayesian approach to an approximation of the classic SIR model.  The approximated SIR model assumes that the incidence at $t+1$, $J(t+1)$ has a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period and $\mathcal{R}$ as the effective reproduction number, which we take here to be $\rr$. In order to estimate $\rr$, we must have some idea about $\gamma$,
\begin{align*}
J(t+1)  \sim Poisson( J(t) \exp \left \{  \gamma (\rr-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidences is
\begin{align*}
  P(\rr | J_0, \dots, J_{t+1}) = \frac{P(J_{t+1} | \rr, J_0, \dots, J_t)P(\rr| J_0, \dots, J_t)}{P(J_0, \dots, J_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.  The advantages of this method are that of the ability to obtain an entire posterior distribution, whereas many other methods are difficult to even find an estimate of the variance.  Disadvantages include strict assumptions about the distributions and computational time required to estimate the relevant parameters.

@@

\cite{wallinga2007generation}, instead, use a moment generating function expression for $\rr$.  With $\omega(t)$ as the serial interval, then
\begin{align*}
\rr^{-1} &= \int_{a=0}^\infty e^{-rt}\omega(t)dt.
\end{align*}
For all these above estimates of $\rr$, we observe the duration of serial intervals in a period of exponential growth.  Deciding when exponential growth occurs and which data points to use to estimate $\rr$ is difficult.  \citeauthor{wallinga2007generation} specify $\omega$ in multiple ways.  Parametrically, they use exponential, normal, and point-wise distributions.  Additionally, they discuss using the the empirical distribution of the serial interval.

@@

%%%%%%%% 5
\subsubsection{Contact Tracing (CT)}
\label{sec:contact_tracing}
A reasonable idea to estimate $\rr$ is to directly observe the number of secondary infections produced by an initial infector.  However, this approach is generally impractical as it is difficult to estimate every single contact an individual may have.   With seriological information, we could conceivably recreate the transmission of the disease throughout a population.  Even in this case it is unlikely, for one, to have a completely susceptible population, which is essential to an unbiased estimate of $\rr$.  Moreover, this would implicitly be conditioning on the initial infector's contacts which may not reflect those of the rest of the population.  We write the calculation for $\rr$ under contact tracing in Equation \ref{eq:r0_contacttracing}, where $X_0$ is a random variable denoting the number of individuals in the population individual 0 infects.  Here $X(0)=N$ indicates that the initial number of susceptibles is $N$, the size of the population,
\begin{align}\label{eq:r0_contacttracing}
\rr = E[ X_0 | X(0) = N].
\end{align}
\cite{eames2003} describe contact tracing as an ``extreme form of targeted control, where the potential next-generation cases are the primary focus.'' That is, one focuses on treating the contacts of an infected individual rather than applying uniform treatment across a population.

@@

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
%\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{graphicx}% http://ctan.org/pkg/graphicx
\usepackage{systeme}
%% tables
\usepackage{booktabs}
\usepackage[table,xcdraw]{xcolor}

@@

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'', $\mathcal{R}_0$, the reproduction number (by convention pronounced ``R-naught''), for nearly a century, has been an elusive quantity for epidemic modelers to estimate \citep{Heesterbeek2002}.  Defined by \citet{anderson1992} is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  $\rr$, in some ways, summarizes an entire outbreak of a disease as this quantity is used to assess if a disease will break out and what percentage of the population needs to be vaccinated to avoid such an outbreak.  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  The primary issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population.

@@

Additionally, we describe  interval estimates for the above methods for estimating $\rr$ in order to better answer the question of whether $\rr > 1$, which implies an outbreak of the disease.  These methods are (1) the delta method, (2) sensitivity analysis, and (3) posterior distribution.

@@

The first set of methods are those of the survival function, which is closely  tied to the history of $\rr$.  The survival function, which originates from the field of demography in the late 1800s, described how many female offspring a woman would produce \citep{dietz1993estimation}.  In demography, we have
\begin{align*}
\rr = \int_0^\infty p(a) \beta(a) da
\end{align*}
where $p(a)$ denotes the probability of a woman surviving to age $a$ and $\beta(a)$ the rate of an individual of age $a$ giving birth to a  girl.   This interpretation also explains the origins of the name of $\rr$, which was imported to the field of epidemiology by MacDonald and Smith \citep{dietz1993estimation}.  Analogously in epidemiology, $p(a)$ is the age of a disease, and $\beta$ is the transmission rate.

This is the chronologically first, and in many ways, the most direct method to estimate $\rr$.  However, this model also seems to be the most difficult to estimate as both $p(a)$ and $\beta (a)$ are random variables.   In general, it is also difficult to estimate these parameters before or even during an ongoing disease outbreak.  As such, $\rr$ can only be estimated after an outbreak, which means it cannot be used to determine ongoing prevention strategies.  However, having a post-disease estimation of $\rr$ may be useful for preventing future outbreaks of the disease and aids in comparing the severity of many diseases.  Another advantage of this method is it intuitive and independent of assumptions of how the disease is transmitted from person to person.    We describe four methods which estimate $\rr$ from the survival function:  the survival function, direct estimation, and direct parameter estimation.

@@

This approach for parameters that are readily available to be estimated, direct parameter estimation, is a `plug-in' estimator approach as described by \cite{lipsitch2003} and \cite{dietz1993estimation}.  This method has an intuitive derivation, similar to that of the survival function.  We assume that we can estimate $\rr$ if we know how many others an individual comes into contact $(k)$, the probability of transmission given contact $(b)$, and how long the disease lasts within an individual $(D)$.  Then $\rr$ is the product of these parameters, 
\begin{align}\label{eq:dpe}
\rr = kbD.
\end{align}
\cite{lipsitch2003} note that directly estimating $\rr$ in such a way may be difficult due to the available data, as in the previous  method.  They instead estimate $\rr$ using the mean serial interval, $\omega$, the time from onset of symptoms in an index case to the onset of symptoms in a subsequent case infected by the index patient in addition to the assumption of exponential growth of infected cases, as a proxy for the duration $D$.  An advantage of this method is that it reduces the problem to three parameters that we can conceivably collect data on.  A disadvantage is thatto estimate $\rr$ alone,  data is only used from the initial outbreak of the disease, so often we have a small sample size, resulting in large confidence sets or credible intervals.

@@

A derivation of $\rr$ similar to the survival function comes from the idea of the von Foerster equations, which were used to calculate the number of cells in a ``deme,'' a constrained area.  The function $Y(t, \tau)$ calculates the number of cells of age $\tau$ at time $t$ \citep{trucco1965}.  In epidemiology, $Y$ represents the number of infections instead of cells. Letting $Y(t, \tau)$ be the number of infections at time $t$ of age $\tau$, the von Foerster equation is
\begin{align*}
  \frac{\partial Y(t, \tau)}{\partial t} +   \frac{\partial Y(t, \tau)}{\partial \tau} = -  \lambda \cdot n(t, a) ,
\end{align*}
where $Y(t,\tau)$ is the number of people at time $t$ who were infected time $\tau$ ago and $\lambda$ is the loss function, which we assume to be 0, that is no deaths.  The right hand side represents the net loss of cells.  This derivation is based on the ``reasoning that the number of cells which are in the age category $(a, a + \Delta a)$ at time $t$, is equal to $n(t,a) \Delta a$.     \cite{fraser2004factors} derived an expression for $\rr$ from the von Foerster equation and the additional boundary equation
\begin{align*}
  Y(t,0) &= \int_0^t \beta( \tau) Y(t, \tau) d \tau,
\end{align*}
and initial conditions $Y(0,0) = Y_0$ and $Y(t, \tau) = 0$ when $t < \tau$.  Then the reproduction number is given by Equation \ref{eq:vfr0},
\begin{align}\label{eq:vfr0}
\rr &= \int_0^\infty \beta(\tau) d\tau,
\end{align}
 where $\beta (\tau)$ represents the infectiousness at time $\tau$ since the infection.    Again, this model is more mathematical in nature than practical as $\beta (\tau)$ is difficult to estimate.  To estimate $\rr$, \cite{fraser2004factors} make an approximation that $Y(t, \tau) = K(\tau) \exp (rt)$ and further adapt the method to include prevention strategies, but the main function to be estimated is $\beta (\tau)$.  Estimating $\beta (\tau)$ has many of the same issues as trying to estimate $b(a)$ in the survival function in that it is difficult to estimate while the disease is ongoing but is also advantageous in that it does not place assumptions on how the disease is transmitted throughout the population.

@@

First introduced by \cite{Kermack700}, a large and the most common set of methods used estimate $\rr$ is that of fitting compartment models (CMs) to disease data.  As described in \cite{daley2001epidemic},  CMs model how objects move from one discrete state to another over time.  These models provide an aggregate model of how diseases may be spread throughout a population.  Within the CM framework, we make four essential assumptions,
\begin{enumerate}
\item The compartments are discrete and have no overlap.
\item The transition of objects into and out of compartments is described by a set of known equations, possibly dependent on unknown parameters.
\item The populations mix homogeneously
  \item The number of objects in each compartment at time $t=0$ is known
  \end{enumerate}

@@

In this section, we discuss $15$ different CM-based methods used to estimate $\rr$.  Different model specifications of CMs yield different mathematical derivations of $\rr$.  That is, for example, an SIR model with non-constant population will yield a different mathematical formula for $\rr$ than the SEIR model (here ``E'' stands for exposed) with non-constant population.  The difference in these two models is that the SEIR model has an extra compartment which in turn effects the estimates of transitions from one compartment to another.  That said, even within the same compartment model framework, we can estimate $\rr$ in a variety of ways using statistical techniques.  In fact, we describe $\xxsir$ methods to estimate $\rr$, within the SIR framework alone!  Most of these methods are recipes to estimate parameters within the SIR model and may be extended to other CM frameworks such as the SEIR model.  The final method described in this section, the next generation operator (NGM) is a general method to derive $\rr$ from any valid CM.  A particular advantage to CMs compared to survival function methods is that it easier to estimate as the data required is typically collected about diseases: namely, incidence counts.

@@

In words, susceptible individuals become infected at a rate that is proportional to the proportion of infected individuals multiplied by $\beta$, the infection rate, and the number of susceptible individuals.  Infected individuals recover at a rate of $\gamma$ multiplied by the number of infected individuals.

An outbreak occurs if the rate of change of infected individuals is positive,
\begin{align*}
  \frac{dY}{dt} &> 0 \\
  \frac{\beta X Y}{N}  - \gamma I &> 0 ,\\
  Y \left ( \beta \frac{X}{N} - \gamma \right ) & > 0
\end{align*}
that is,  the rate of new infections is greater than the rate of recovery.  So as long as the number of susceptibles is large, $\frac{X}{N} \approx 1$, then an outbreak will occur if $\rr >1$,
\begin{align}\label{eq:deriv_sir}
  \rr = \frac{\beta}{\gamma} > 1.
  \end{align}
Ideally, we have access data of the number of susceptibles, infecteds, and recovered individuals at different points over time.  From this we would like to estimate the $\beta$ and $\gamma$ parameters and ultimately $\rr$.

@@

\begin{align*}
(\hat{\beta}, \hat{\gamma} )&=\text{argmin}_{\beta, \gamma} \sum_{t} \left [ \left (Y_{obs}(t) - Y(t)\right )^2 + \left ( X_{obs}(t) - X(t) \right )^2 \right ]
\end{align*}
Then  the estimate for $\rr$ is given by Equation \ref{eq:sirls},
\begin{align}\label{eq:sirls}
  \hat{R}_0= \frac{\hat{\beta}}{\hat{\gamma}}.
  \end{align}
We can use gradient descent or another optimizer to find estimates for these parameters.  If we assume $Y= f(t) + \epsilon_t$ with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$ where $f(t) = \int_0^t \frac{dY}{dt} dt$, then this this method is equivalent to finding the maximum likelihood estimator (MLE) of $\beta$ and $\gamma$.  However, we note that the the $\epsilon_t$ are certainly not independent from one another and so valid inference is still an issue in this framework.  Even if it were correct, ratio estimatiors typically result in large, uninformative confidence intervals.   The advantage is that this is a very straightforward way to estimate $\rr$ from the SIR model.

@@

\begin{align*}
  \left \{
  \begin{array}{cl}
    \frac{dX}{dt} &= - \rr \gamma Y \frac{X}{N}\\
    \frac{dY}{dt} &=  \rr \gamma Y \frac{X}{N}  - \gamma Y \\
    \frac{dZ}{dt} &=  - \gamma Y 
  \end{array}
  \right .
  \end{align*}
We find
\begin{align*}
(\hat{R}_0, \hat{\gamma} ) &= \text{argmin}_{\beta, \gamma} \sum_{t} \left [ \left (Y_{obs}(t) - Y(t)\right )^2 + \left ( X_{obs}(t) - X(t) \right )^2 \right ]
\end{align*}
We use the \(\rr\) directly from the above estimation process, which again be done with gradient descent or another optimization process.  Again, we cannot assume $\rr$ has a normal distribution as the errors are correlated.  This method is better than SIR LS for estimating $\rr$ as there is no ratio.

@@

Specifically, we fit two linear polynomials in \(t\) with degree \(K= 10\) to \(X_{obs}\)
and \(Y_{obs}\) using least squares to find the coefficients (\(\hat{x}_k\),
\(\hat{y}_k\)),
\begin{align*}
\hat{X}(t) &= \sum_{k=0}^K \hat{x}_k t^k\\
{\hat{Y}}(t) &= \sum_{k=0}^K \hat{y}_k t^k
\end{align*}
Then, we estimate the derivatives as
\begin{align*}
\hat{X}^\prime(t) &= \sum_{k=1}^K k \hat{x}_k t^{k-1}\\
\hat{Y}^\prime(t) &= \sum_{k=0}^K k \hat{y}_k t^{k-1}
\end{align*}

Following,  \(\rr\) is derived from the ODEs in Equation \eqref{eq::sir},
\begin{align}
  - \frac{X^\prime}{Z^\prime}&= \rr \frac{X}{N}\\
\rr &=       \frac{X}{ X + Y} \cdot \frac{N}{X}\\
  \hat{\rr} &= \frac{\hat{X}^\prime(0)}{ \hat{X}^\prime(0) + \hat{Y}^\prime(0)} \cdot \frac{N}{X(0)}.
  \end{align}
Like in the previous cases, we are assuming the true models are
\begin{align*}
X(t) &= \sum_{k=0}^K x_k t^k + \epsilon_t\\
  {Y}(t) &= \sum_{k=0}^K y_k t^k + \epsilon_t,
\end{align*}
with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$.  Here, $K=10$ is rather arbitrary and should be selected using some criterion such as AIC in addition with the knowledge of how many data points are available to the user.  Besides optionally deciding on the degree of polynomials to fit, this model is simple to implement and gives very comparable results to using least squares with the SIR model.  The time $t=0$ is used as this is best capture the initial outbreak.  Again, we cannot use this model for inference, as our error assumption is violated.  Another disadvantage is that we are using the first time point $t=0$ to estimate $\rr$.

@@

\subsubsection{Max of Data (SIR Max)}\label{max-of-data}
The I compartment of the SIR model has one peak.    We note that \(Y^\prime(t) = \frac{\beta XY}{N} - \gamma Y = 0\) when \(\rr = \frac{\beta}{\gamma} = \frac{N(t)}{X(t)}\). The max
of \(Y\) occurs when \(Y^\prime = 0\) and hence
\[\rr = \frac{N}{X_{obs}(t^*)},\] where
\(t^* = \text{arg max}_{t} Y_{obs}(t)\).   A major disadvantage to this method is that it only uses one sample, making it a highly variable statistic.  An advantage of this method is that it places no distributional assumptions on any parameter.

@@

Here we assume that the time step is small enough to approximate the incidence.  The advantage of this method are that incidence data is generally readily available as is prevalence data for certain diseases such as HIV.  However, as one is required to have prior knowledge about $\gamma$, it may be easier to directly estimate $\rr$ with one of the many other methods described that does not require a prior knowledge about $\gamma$.  Again, we are using only one time point to estimate $\rr$.

@@

In this framework, the number of susceptibles at the current time, $X_{t+1}$, has a binomial distribution based on the contacts with the current number of infected, $Y_t$ and the current number of susceptibles.  That is $X_{t+1} \sim \text{Binom}(X_t, \alpha^{Y_t}$), where $\alpha$ is the probability of avoiding infection from an infective.  \cite{barbour2004} report that the reproduction number is thus,
\begin{align}\label{eq:r0-mc}
\rr &= \log \left ( \frac{1}{1-\alpha}\right ).
\end{align}

@@

Described by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, the sequential Bayes method is a Bayesian approach to an approximation of the classic SIR model.  The approximated SIR model assumes that the incidence at $t+1$, $J(t+1)$ has a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period and $\mathcal{R}$ as the effective reproduction number, which we take here to be $\rr$. In order to estimate $\rr$, we must have some idea about $\gamma$,
\begin{align*}
J(t+1)  \sim Poisson( J(t) \exp \left \{  \gamma (\rr-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidences is
\begin{align*}
  P(\rr | J_0, \dots, J_{t+1}) = \frac{P(J_{t+1} | \rr, J_0, \dots, J_t)P(\rr| J_0, \dots, J_t)}{P(J_0, \dots, J_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.  The advantages of this method are that of the ability to obtain an entire posterior distribution, whereas many other methods are difficult to even find an estimate of the variance.  Disadvantages include strict assumptions about the distributions and computational time required.

@@

\cite{wallinga2007generation}, instead, use a moment generating function expression for $\rr$.  With $\omega(t)$ as the serial interval, then
\begin{align*}
\rr^{-1} &= \int_{a=0}^\infty e^{-rt}\omega(t)dt.
\end{align*}
For all these above estimations of $\rr$, we observe the duration of serial intervals in a period of exponential growth.  Deciding when exponential growth occurs and which data points to use to estimate $\rr$ is difficult.  \citeauthor{wallinga2007generation} specify $\omega$ in multiple ways.  Parametrically, they use exponential, normal, and point-wise distributions.  Additionally, they discuss using the the empirical distribution of the serial interval.

@@

%%%%%%%% 5
\subsubsection{Contact Tracing (CT)}
\label{sec:contact_tracing}
A reasonable idea to estimate $\rr$ is to directly observe the number of secondary infections produced by an initial infector.  However, this approach is generally impractical as it is difficult to estimate every single contact an individual may have.   With seriological information, we could conceivably recreate the transmission of the disease throughout a population.  Even in this case it is unlikely, for one, to have a completely susceptible population, which is essential to an unbiased estimate of $\rr$.  Moreover, this would implicitly be conditioning on the initial infector's contacts which may not reflect those of the rest of the population.  We write the calculation for $\rr$ under contact tracing in Equation \ref{eq:r0_contacttracing}, where $X_0$ is a random variable denoting the number of individuals in the population individual 0 infects.  Here $S_0=N$ indicates that the initial number of susceptibles is $N$, the size of the population,
\begin{align}\label{eq:r0_contacttracing}
\rr = E[ X_0 | S_0 = N].
\end{align}
\cite{eames2003} describe contact tracing as an ``extreme form of targeted control, where the potential next-generation cases are the primary focus,'' that is, one focuses on treating the contacts of an infected individual rather than applying uniform treatment across a population.

@@

\subsubsection{Sensitivity Analysis}\label{sensitivity-analysis}
